---
title: GT discrepencies
output:
  html_document:
    df_print: paged
    code_folding: show
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "html_output"
    )
  })
---
```{r include = FALSE}
library(httpgd)
library(VariantAnnotation)
library(data.table)
library(parallel)
library(DT)
library(htmltools)
library(ggplot2)
```

## GT imbalances {.tabset .tabset-fade .tabset-pills}

### Intro

#### Data.table linking samples to groups

##### Rational:
Let's create a table where the samples in the VCF are associated with a group representing a set
of clones carrying the same CRISPR edit type.

##### Format:
The iSCORE-PD_design.csv file is a comma seperated text file starting with a header and with one cell line per line with the following
columns:

- samples: Sample ID found in VCF header of the joint genotyping output file
- group: Group ID linking the sample in group
- meta: Additional group relation. Not used anymore

```{r}
samplesDT <- setDT(read.csv("./supporting_files/iSCORE-PD_design.csv"))
datatable(samplesDT)
```

### GT Analysis

#### Rational

Many SNPs in the sampeles have more than one GT, more than the number of unique SNPs. What's going on?

#### Fraction of samples carying the same genotypes

Ok, Let's look at chromosome 1 only 
```{r}
vcf <- readVcf(vcf.files[1])

groups <- samplesDT[!group %in% c("parental", "EWT"), unique(group), ]

## Get the sample names of the group under analysis
currUniverse <- samplesDT[-grep("EWT", group), samples]

asSNPs <- apply(geno(vcf)$GT[, currUniverse], 1, function(x) !all(x %in% c("./.", "0/0", "./0", "0/.")))

vcf_with_snps <- vcf[asSNPs & qual(vcf) >= 30]

asSameGT <- apply(geno(vcf_with_snps)$GT[, currUniverse], 1, function(x) length(unique(x)) == 1)

## Fraction of SNPs that have the same GT over all the samples
sum(asSameGT) / length(asSameGT)
```

#### Franction of SNPs where all genotypes have an uncalled allele
```{r}
geno <- geno(vcf_with_snps[!asSameGT])$GT[, currUniverse]

lvls <- combn(0:4, 2, paste, collapse = "/")
lvls <- c(lvls, sapply(0:4, function(x) paste(rep(x, 2), collapse = "/")))

unique_geno <- apply(geno, 1, function(x) unique(x[-grep("\\.", x, )]))

## Franction of SNPs where all genotypes have an uncalled allele
sum(!elementNROWS(unique_geno) > 1) / length(unique_geno)
```

#### Number of unique GT that have calls
```{r}
unique_geno <- unique_geno[elementNROWS(unique_geno) > 1]
length(unique_geno)
```

#### Fraction of GT transition from Het->Homo (or vice versa)

Turns out that a large fraction of the discordant GT are transition between Het and Homo.

```{r warning=FALSE}
dt1 <- data.table(
  snp = names(unique_geno),
  gts = sapply(unique_geno, function(x) paste(factor(x, levels = lvls), collapse = "|"))
)

dt <- data.table(table(sapply(unique_geno, function(x) paste(factor(x, levels = lvls), collapse = "|"))))
setnames(dt, "V1", "gts")
dt[, gtCount := elementNROWS(strsplit(gts, "\\|"))]
dt[, alleleCount := unlist(lapply(strsplit(gts, "\\||/"), function(x) length(unique(x))))]
dt[, homo2het := "No"]
dt[!(alleleCount > 2 | gtCount > 2), homo2het := lapply(strsplit(gts, "\\|"), function(y) {
  yss <- lapply(strsplit(y, "/"), unique)
  if (any(yss[[1]] %in% yss[[2]] | yss[[2]] %in% yss[[1]])) "yes"
})]

dt[order(factor(homo2het, levels = c("yes", "no")))]

dt[homo2het == "yes", sum(N)] / dt[, sum(N)]
```

Most likely, the source of these transitions are homologous DNA repair, where during replication, a break hapened
and is being recessed then repaired by the homolgous chromatid. Any Het SNPs in within the recessed region will be
converted to the reciprocal allele, converting the site from Het to Homo

### Dist Analysis

#### Distribution of distances between any 2 SNPs

If the hypothesis is true, there's a greather likely hood to transition SNPs from Het to Homo that are closer to each others.
Let's look at that.

```{r}
homo_2_het_snps <- dt1[gts %in% dt[homo2het == "yes", gts], snp]
start1 <- start(vcf[homo_2_het_snps])
start2 <- c(start1[-1], seqlengths(vcf)[unique(seqnames(vcf))])

df <- data.frame(
  dist2next = start2 - start1,
  type = "homo2het"
)

not_homo_2_het_snps <- dt1[!gts %in% dt[homo2het == "yes", gts], snp]
start3 <- start(vcf[not_homo_2_het_snps])
start4 <- c(start3[-1], seqlengths(vcf)[unique(seqnames(vcf))])

df <- rbind(df, data.frame(
  dist2next = start4 - start3,
  type = "notHomo2het"
))

p <- ggplot(df, aes(log10(dist2next), color = type)) +
  geom_density()

print(p)
```

### Read Depth

#### Computing the Read Depth distrubtion for Major vs Minor Genotypes

Let's take the SNPs that have samples with genotyp transitions from het to homo and look at the read depth coverage from the sample correspoding
to the major GT (ie samples from the group with most of one of the two GT) compare to the samples with the minor GT (ie sample
from the group wiht the least of one of the two GT). Also, as a control, let's sample the compute the read coverage from SNPs where all 
the genotypes are identical (well do some sampling here to reduce the size of the finale data frame).

```{r}
df <- do.call(rbind, mclapply(dt[homo2het == "yes" & N > 10, gts], function(currCase) {
  gts <- strsplit(currCase, "\\|")[[1]]
  currVcf <- vcf[dt1[gts == currCase, snp], currUniverse]

  majorDP <- unlist(sapply(seq_len(length(currVcf)), function(i) {
    major <- which.max(c(sum(geno(currVcf)$GT[i, ] == gts[1]), sum(geno(currVcf)$GT[i, ] == gts[2])))
    f <- geno(currVcf)$GT[i, ] == gts[major]
    geno(currVcf)$DP[i, f]
  }))

  minorDP <- unlist(sapply(seq_len(length(currVcf)), function(i) {
    major <- which.min(c(sum(geno(currVcf)$GT[i, ] == gts[1]), sum(geno(currVcf)$GT[i, ] == gts[2])))
    f <- geno(currVcf)$GT[i, ] == gts[major]
    geno(currVcf)$DP[i, f]
  }))

  f <- apply(geno(vcf[, currUniverse])$GT, 1, function(x) length(unique(x)) == 1 && unique(x) %in% gts)
  uniqueDP <- as.vector(geno(vcf[f][sample(sum(f), length(currVcf))])$DP)

  df <- data.frame(
    DP = c(majorDP, minorDP, uniqueDP),
    sample = c(
      rep("Major GT", length(majorDP)),
      rep("Minor GT", length(minorDP)),
      rep("Unique GT", length(uniqueDP))
    )
  )
}, mc.preschedule = FALSE, mc.cores = detectCores()))

p <- ggplot(df, aes(DP, color = sample)) +
  geom_density() +
  xlim(0, 60) +
  labs(
    title = "Distributions of squencing depth (deepVariant DP) between samples from major, minor, or common GT",
    x = "Read Deapth"
  )

print(p)
tapply(df$DP, df$sample, summary)
```

#### Computing the Allelic Read Depth Ratio for Major vs Minor Genotypes

Looking at same SNPs as above but computing the ration of read depth for one allele compare to the other allele


```{r}
df <- do.call(rbind, mclapply(dt[homo2het == "yes" & N > 10, gts], function(currCase) {
  gts <- strsplit(currCase, "\\|")[[1]]
  currVcf <- vcf[dt1[gts == currCase, snp], currUniverse]

  majorAD <- unlist(sapply(seq_len(length(currVcf)), function(i) {
    major <- which.max(c(sum(geno(currVcf)$GT[i, ] == gts[1]), sum(geno(currVcf)$GT[i, ] == gts[2])))
    f <- geno(currVcf)$GT[i, ] == gts[major]
    AD <- geno(currVcf)$AD[i, f]
    sapply(AD, function(x) x[1] / sum(x))
  }))

  minorAD <- unlist(sapply(seq_len(length(currVcf)), function(i) {
    minor <- which.min(c(sum(geno(currVcf)$GT[i, ] == gts[1]), sum(geno(currVcf)$GT[i, ] == gts[2])))
    f <- geno(currVcf)$GT[i, ] == gts[minor]
    AD <- geno(currVcf)$AD[i, f]
    sapply(AD, function(x) x[1] / sum(x))
  }))

  f <- apply(geno(vcf[, currUniverse])$GT, 1, function(x) length(unique(x)) == 1 && unique(x) %in% gts)
  sampleVcf <- vcf[f][sample(sum(f), length(currVcf))]

  uniqueAD <- unlist(sapply(seq_len(length(sampleVcf)), function(i) {
    AD <- geno(sampleVcf)$AD[i, ]
    sapply(AD, function(x) x[1] / sum(x))
  }))

  df <- data.frame(
    AD = c(majorAD, minorAD, uniqueAD),
    sample = c(
      rep("Major AD", length(majorAD)),
      rep("Minor AD", length(minorAD)),
      rep("Unique AD", length(uniqueAD))
    )
  )
}, mc.preschedule = FALSE, mc.cores = detectCores()))

ggplot(df, aes(AD)) +
  geom_histogram() +
  facet_grid(rows = vars(sample), scales = "free") +
  labs(
    title = "Distributions of fraction of Allele Depth (deepVariant AD)\n between samples from major, minor, or common GT", # nolint: line_length_linter.
    x = "Read Deapth"
  )


tapply(df$AD, df$sample, summary)
```

### sessionInfo
```{r}
sessionInfo()
```